FastSRM: A fast, memory efficient and identifiable implementation of the shared response model

### 2021-11-30 paper/main.tex
A pass on the paper

\begin{abstract}
  The shared response model (SRM) provides a simple but effective framework to analyse
  fMRI data of subjects exposed to naturalistic stimuli.
  %
  However when the number
  of subjects or runs is large,
  % \bt{True, but we the experiments don't make it so explicit ?}
  fitting the model requires a large amount of memory
and computational power, which limits its use in practice.
%
Furthermore, SRM is
not identifiable which makes the shared response difficult to interpret.

In this work, we implement an identifiable version of SRM and show on real data
that it improves the stability of the recovered shared response.
%
We then introduce the FastSRM that relies on a dimension reduction step that we
prove to be optimal in the sense that working with reduced data does not induce any change in the
algorithm trajectory.
%
We show experimentally using synthetic and real fMRI data
that FastSRM is considerably faster and more memory efficient
than competitive implementations.

The experiments performed in the paper are \emph{fully} reproducible: our
code available at \url{https://github.com/hugorichard/FastSRM} downloads the
data, runs the experiments and plot the figures.
%

@@

\section{Introduction}
%\bt{We need to start with a more application-oriented intro}
%
Complex stimulations such as movie-watching, story or music listening are more
and more popular in the neuro-scientific community. Indeed such naturalistic
paradigms are unconstrained from behavioral manipulations and thus, more
ecological with respect to real-world conditions.
%
However and in contrast to more traditional designs, these so-called
naturalistic stimuli do not come with a design matrix that quantifies the
features of interest in the stimuli.
%
Without a design matrix, models such as the general linear
model~\cite{poline2012general} cannot be used. Although some works have used
natural annotations~\cite{huth2012continuous} or deep neural
networks~\cite{gucclu2017increasingly}~\cite{richard2018optimizing} to create an
estimate of the design matrix of naturalistic
stimuli it is by no means a simple task.

@@

In this work, we introduce FastSRM, an identifiable model that uses an optimal
atlas-based decomposition to speed up the computations with provably no loss of
performance.
%
FastSRM gains identifiability by imposing that the covariance of the shared
components is diagonal (see ~\cite{richard2020modeling} Appendix D for a proof).
%
The gains in speed and memory usage are based on the intuition that a compressed
representation of the input should be good enough to find a suitable estimate of
the shared components as they live in reduced space.
%
It turns out that there
exists an optimal atlas that reduces the data in a such a way that the obtained
shared components (and other parameters) are the same as if they were computed
on full data.
%
We show that FastSRM brings several practical benefits: on real data, it produces more stable
estimates of the shared components than its non-identifiable counter-part and 
 it is much faster and more memory
efficient than other implementations that do not make use of an optimal atlas.
% on synthetic as well as on real data,

Our python implementation is available at
\url{https://github.com/hugorichard/FastSRM}.
%
All our experiments are fully reproducible.
%
Scripts to download the data, run the experiments and plot the figures are
included.
%
A continuous integration pipeline runs the tests
automatically when any change to the core algorithm is made.
%

\section{Background: the shared response model (SRM)}
\label{sec:srm:review}
The shared response model~\cite{chen2015reduced} is a multi-view latent factor
model.
%
The data $\xb_1 \dots \xb_m$ are modeled as random vectors following the model:
\begin{align}
 &\xb_i = A_i \sbb + \nb_i \\
  &A_i^{\top}A_i = I_p
  \label{eq:model:srm}
\end{align}
where $\xb_i \in \RR^v$ is the data of view $i$, $A_i \in \RR^{p \times v}$ is the
mixing matrix of view $i$, $\nb_i$ is the noise of view $i$ and $\sbb \in \RR^p$ are the
shared components referred to as the \emph{shared response} in fMRI
applications.
%

@@

The conceptual figure~\ref{fig:srm:conceptual_figure} illustrates an 
application of the shared response model to fMRI data.
%
The mixing
matrices are spatial topographies specific to each subjects while the shared
components give the common timecourses.
%
In~\cite{chen2015reduced, anderson2016enabling}, two versions of the shared
response model are introduced which we now present.
%

@@

The model is optimized by maximizing the log-likelihood, assuming that
the noise is Gaussian distributed.
%
The likelihood is then given by: $p(\xb) = \prod_i \Ncal(\xb_i; A_i \sbb, \sigma^2 I)$ and
therefore the empirical expected negative log-likelihood is given up to a constant independent of
$A_i$ and $S$ by:
\begin{align}
  \loss = \frac1{n} \sum_i \|A_i S - X_i \|^2 = \frac1{n} \big(\| S \|^2 -2 \dotp{ A_i S}{ X_i } + \| X_i \|^2 \big)
  \label{eq:detsrmloss}
\end{align}
$\loss$ is optimized by performing alternate minimization on $(A_1 \dots A_m)$
and $S$.
%
Note that the hyper-parameter $\sigma$ does not have an influence on
the loss and can therefore be ignored.

@@

From \eqref{eq:detsrmloss}, minimizing $\loss$ with respect to $A_i$ is
equivalent to maximizing $\dotp{ A_i}{ X_i S^{\top} }$ and therefore we
have:
\begin{equation}
  A_i \leftarrow  \Pcal(\frac1{n}X_i S^{\top})
  \label{eq:detsrm:Aiupdate}
\end{equation}
where $\Pcal$ is the projection on the Stiefel manifold: $\Pcal(M) = M
(M^{\top}M)^{-\frac12}$.
%

The complexity of Deterministic SRM is in $\bigO(\mathrm{n_{iter}} mpvn)$ where
$n$ is the number of samples and $\mathrm{n_{iter}}$ the number of iterations.
%
We monitor the convergence by looking at the $\ell_{\infty}$ norm of the
gradient.
%
Note that we can monitor the gradient without any increase in complexity.
%
Indeed, after the updates with respect to each mixing matrix have been
carried out, only the gradient with respect to $S$ remains: $\sum_i
(S - A_i^{\top}\xb_i)$.
%
The algorithm is stopped when the
gradient falls below a chosen tolerance.
%

@@

In~\cite{chen2015reduced} and~\cite{anderson2016enabling}, $\Sigma_s$ is only assumed to be definite positive.
%
As already highlighted in introduction, this causes the model to be
unidentifiable (see~\cite{richard2020modeling} Appendix D for a proof). Enforcing a diagonal $\Sigma_s$ ensures identifiability (provided the diagonal values are different).
%
So we assume here that $\Sigma_s$ is diagonal (and refer the interested reader
to~\cite{chen2015reduced} and~\cite{anderson2016enabling} for the original formulation of Probabilistic SRM
without the diagonal constraint)
%

@@

\subsection{Reducing the computational burden by use of atlases}
SRM algorithms use different set of parameters $\theta$ to represent the data.
%
In deterministic SRM $\theta = \left((A_i)_{i=1}^m, S \right)$ where $(A_i)_{i=1}^m$ are the
mixing matrices and $S$ is the shared response, while in probabilistic SRM $\theta
= \left( (A_i)_{i=1}^m, \Sigma_s, (\sigma_i)_{i=1}^m \right) $ where $(A_i)_{i=1}^m$ are the
mixing matrices, $(\sigma_i)_{i=1}^m$ the noise levels and $\Sigma_s$ the
components variance.
%

@@

In FastSRM we consider a set of view-specific atlases $U_i \in \RR^{v \times r}$ such that
$U_i^{\top}U_i = I_r$ where $r$ is the number of regions in the atlas.
%
Data are reduced using $\zb_i = U_i^{\top} \xb_i$ and an SRM algorithm is applied
on data $\zb_i$ yielding parameters $\theta'$.
%
The figure~\ref{fig:srm:conceptual} illustrates this process.
%

@@

\subsection{An optimal atlas}
\label{sec:optimal_atlas}
In principle, FastSRM can be used with any atlas.
%
However, in general, working with reduced data induces a loss of information
that can be minimized if the atlas is carefully chosen.
%
In any case, there is little hope to recover the parameters that would have been
obtained from SRM from the parameters of FastSRM.
%
Yet, we show that there exists an optimal atlas in the
sense that SRM and FastSRM yield the same results.

Let us consider $\xb_i = U_{\xb_i} \zb_i$
%
% \bt{But any fill-rank orthogonal basis would work equally well (i.e. any
%   orthogonal transform of the PCA basis)}
%
a PCA of $\xb_i$ with the maximum
number of components.
%
As the number of
samples $n$ is lower than the number of features, $U_{\xb_i} \in \RR^{v \times
  n}$ and $\zb_i \in \RR^{n}$.
%
We also have $U_{\xb_i}^{\top}U_{\xb_i} = I$.
%
Therefore $U_{\xb_i}$ constitutes a possible choice of subject specific atlas.
%
As the next property shows, $U_{\xb_i}$ is an optimal atlas for deterministic
FastSRM. Note that this optimal atlas is not unique: any orthogonal matrix with
the same span as $X_i$ would also work.
%

@@

In the case of probabilistic SRM we can obtain very similar results.
%
However the
algorithm applied on reduced data need to be slightly modified.
%

We have the following result:
\begin{proposition}[Optimal atlas for probabilistic FastSRM]
  Let $(A_i)_i, \sigma_i, \Sigma_s$ be the solution obtained by probabilistic SRM on data
  $\xb_i$ and $(A_i')_i, \sigma_i', \Sigma_s'$ the solution obtained by ProbSRM on
  data $\zb_i = U_{\xb_i}^{\top} \xb_i$ but where updates:
  \begin{align}
    \sigma_i^2 \leftarrow \frac1{t} (\EE[\| \zb_i - A_i \EE[\sbb|\zb]\|^2] + \| \diag(\VV[\sbb | \zb]) \|^2)
  \end{align}
  are replaced by updates
  \begin{align}
    \sigma_i^2 \leftarrow \frac1{v} (\EE[\| \zb_i - A_i \EE[\sbb|\zb]\|^2] + \| \diag(\VV[\sbb | \zb]) \|^2)
  \end{align}
  .
  %
  Then $A_i = U_{\xb_i}A_i'$, $\sigma_i =
  \sigma_i'$ and $\Sigma_s = \Sigma_s'$.
  %

@@

A complexity analysis shows that finding an optimal atlas becomes the limiting
step of the pipeline.
%
Even with fast implementations, the subject specific PCA
is costly.
%
However FastSRM only works on $\zb_i$ so we do not need to know the value of
$U_{\xb_i}$.
%
In practice, we observe data $X_i \in \RR^{v \times n}$ and we want to get $Z_i
\in \RR^{n \times n}$ such that $X_i = U_{\xb_i} Z_i$.
%
This can be done by performing an
SVD of $X_i^{\top} X_i$ yielding $X_i^{\top}X_i= V_i D_i V_i^{\top}$ and setting
$Z_i = D_i^{\frac12} V_i^{\top}$.
%
Although computing the product $X_i^{\top} X_i$ has time complexity
$\bigO(vt^2)$ (the same as a PCA) the constant in the $\bigO$ is exactly one so
this operation costs a lot less than the PCA
on full data.
%
When estimates of the mixing matrices are needed, they can be obtained by
applying equation~\eqref{eq:fastdetsrm:Aiupdate} in the deterministic SRM case and
equation~\eqref{eq:fastprobsrm:Aiupdate} in the probabilistic SRM case which only costs
$\bigO(mvp^2)$.
%
In practice the cost of the matrix products $X_i^{\top} X_i$ is often still the
limiting step of the pipeline (this depends on the number of iterations) but as
we show next, it is much more efficient than performing SRM on
the full data.
%
Note that if memory allows it, these matrix products can be
computed in parallel.
%

\section{Related work}
The implementation of SRM proposed in~\cite{chen2015reduced} is quadratic in the
number of features which prevent its application to full brain data.
In~\cite{anderson2016enabling}, the authors address this issue by using the
inversion lemma to remove the quadratic dependence. Their implementation is now the most widely used one.
In our work, we propose to reduce further the computation time and memory usage
compared to the implementation in~\cite{anderson2016enabling} by the use of
optimal atlases.

@@

We benchmark deterministic SRM, probabilistic SRM and their FastSRM counterparts in terms of fitting time and performance.
%
Note that in this section, the identifiable implementation of deterministic SRM
and probabilistic SRM described in section~\ref{sec:probabilisticsrm}
and~\ref{sec:deterministicsrm} is used. 
%
Algorithms are designated by the atlas they use and therefore SRM algorithms are refered to
as \emph{None} because no atlas is used and FastSRM algorithms will have the
label \emph{Optimal}.
%
Note that it would be possible to use FastSRM with sub-optimal
atlases (there exists a wide variety of brain atlases
~\cite{schaefer2017local, bellec2010multi, mensch2018extracting}) but
using them does not brain any guarantee that the performance is the same as SRM.
%

@@

\subsection{Identifiability increases stability}
Assuming that the data follow the model, identifiability ensures that the
problem is well posed since the parameters of the model are not ambiguous. Note however that SRM
stays non-convex so there is no guarantee that the optimal solution is found.
%
% \bt{I find this a bit ambiguous: you give te impression that you gurantee that the optimal solution is used. Rephrase.}
%
In real fMRI data, the model cannot be expected to hold perfectly, but we can
hope for greater stability in the parameters recovered than if an unidentifiable
model is used.
%

To measure the stability of the common components obtained from the Sherlock
dataset, we divide the subjects into two roughly equal groups and extract the
common components in each group.
%
The components are then matched using the Hungarian algorithm and the stability
measure is obtained from the average correlation of the matched components.
%
The procedure is repeated 9 times using the Brainiak implementation (that is not
identifiable) and our
FastSRM implementation.
%
We plot the results as a scatter plot in Figure~\ref{exp:identifiability}, where, for each repetition , the x-axis indicates the stability measure obtained with Brainiak's implementation
(not identifiable since the shared components covariance is unconstrained) and the y-axis the stability measure
obtained with FastSRM (identifiable since the shared components covariance is diagonal).
%
We see that introducing a diagonal source covariance improves stability.
%

@@

We show on synthetic and real data that FastSRM is much faster and more memory
efficient that implementations not using an optimal atlas.
Furthermore, FastSRM is identifiable. On real data, we show that compared to
Brainiak's implementation (which is not identifiable), FastSRM provides more
stable estimates of the shared components.

%
FastSRM inherits from all the weaknesses of SRM including the fact that
mixing matrices are assumed to be orthogonal which is rather unrealistic. It
remains to be seen whether dimension reduction and learning of the shared
components could be done jointly and efficiently without assuming orthogonal
mixing matrices.
%
Our optimal atlas approach shows that SRM-like methods do not need the
full data to provide accurate estimates of shared components. We believe such
insights may guide the design of future methods.
%

Lastly, FastSRM is useful when the number of features is much larger than the
number of samples. This setting is common in fMRI but less common in MEG/EEG where
the number of samples is usually much larger than the number of features. Even
in this case, it is possible to considerably speed up SRM. Indeed, SRM only
depends on second order statistics.
These statistics can be pre-computed making the rest of the
pipeline insensitive to the number of samples in the dataset. An implementation
of this simple idea would make SRM easily applicable to other modalities.
We live it to future work.
% \bt{Add a paragraph on implementation, sofwtare quality, documentation, examples, availability...}
% \bt{Discuss the use for other brain imaging modelities (MEEG) if relevant ?}

>>>>

\begin{abstract}
  The shared response model (SRM) provides a simple but effective framework to analyse
  fMRI data of subjects exposed to naturalistic stimuli.
  %
  However when the number
  of subjects or runs is large,
  % \bt{True, but we the experiments don't make it so explicit ?}
  fitting the model requires a large amount of memory
and computational power, which limits its use in practice.
%
Furthermore, SRM is
not identifiable, which makes the shared response difficult to interpret.

In this work, we implement an identifiable version of SRM and show on real data
that it improves the stability of the recovered shared response.
%
We then introduce the FastSRM that relies on a dimension reduction step that we
prove to yield the same solution as the original algorithm.
% be optimal in the sense that working with reduced data does not induce any change in the
% algorithm trajectory.
%
We show experimentally using synthetic and real fMRI data
that FastSRM is considerably faster and more memory efficient
than current implementations.

The experiments performed in the paper are \emph{fully} reproducible: our
code available at \url{https://github.com/hugorichard/FastSRM} allows to download the
data, run the experiments and plot the figures.
%

@@

\section{Introduction}
%\bt{We need to start with a more application-oriented intro}
%
Complex stimulations such as movie-watching, story or music listening are more
and more popular in the neuro-scientific community. Indeed such naturalistic
paradigms are unconstrained from behavioral manipulations and thus, more
ecological with respect to real-world conditions.
%
However and in contrast to more traditional designs, these so-called
naturalistic stimuli do not come with a design matrix that quantifies the
features of interest in the stimuli.
%
Without a design matrix, models such as the general linear
model~\cite{poline2012general} cannot be used. Although some works have used
natural annotations~\cite{huth2012continuous}, see also \url{https://neuroscout.org} or deep neural
networks~\cite{gucclu2017increasingly}~\cite{richard2018optimizing} to create an
estimate of the design matrix of naturalistic
stimuli it is by no means a simple task.

@@

In this work, we introduce FastSRM, an identifiable model that uses an optimal
atlas-based decomposition to speed up the computations with provably no loss of
performance.
%
FastSRM gains identifiability by imposing that the covariance of the shared
components is diagonal (see ~\cite{richard2020modeling} Appendix D for a proof).
%
The gains in speed and memory usage are based on the observation that a low-dimensional, yet distance-preserving
representation of the images yields the same result as the full data.
%
Such a representation can be interpreted as an initial spatial  decomposition of the data.
% that reduces the data in a such a way that the obtained
% shared components (and other parameters) are the same as if they were computed
% on full data.
%
We show that FastSRM brings several practical benefits: on real data, it produces more stable
estimates of the shared components than its non-identifiable counterpart and 
 it is much faster and more memory
efficient than other implementations that do not make use of an optimal atlas.
% on synthetic as well as on real data,

A Python implementation is available at
\url{https://github.com/hugorichard/FastSRM}.
%
All our experiments are fully reproducible.
%
Scripts to download the data, run the experiments and plot the figures are
included.
%
A continuous integration pipeline runs the tests
automatically when any change to the core algorithm is made.
%

\section{Background: the shared response model (SRM)}
\label{sec:srm:review}
The shared response model~\cite{chen2015reduced} is a multi-view latent factor
model.
%
The data $\xb_1 \dots \xb_m$ are modeled as random vectors following the model:
\begin{align}
 &\xb_i = A_i \sbb + \nb_i \\
  &A_i^{\top}A_i = I_p
  \label{eq:model:srm}
\end{align}
where $\xb_i \in \RR^v$ is the data of subject $i$, called view $i$, $A_i \in \RR^{p \times v}$ is the
mixing matrix of view $i$, $\nb_i$ is the noise of view $i$ and $\sbb \in \RR^p$ are the
shared components referred to as the \emph{shared response} in fMRI
applications.
%
$p$ is the number of time frames, $v$ is the number of voxels and $m$
is the number of subjects.

@@

The conceptual figure~\ref{fig:srm:conceptual_figure} illustrates an 
application of the shared response model to fMRI data.
%
The mixing matrices are spatial topographies specific to each subject
while the shared components represent the common timecourses.
%
In~\cite{chen2015reduced, anderson2016enabling}, two versions of the
shared response model are introduced, namely the deterministic and
probabilistic models.

@@

The model is optimized by maximizing the log-likelihood, assuming that
the noise is Gaussian distributed.
%
The likelihood is then given by: $p(\xb) = \prod_i \Ncal(\xb_i; A_i \sbb, \sigma^2 I)$ and
therefore the empirical expected \textbf{XXX why expected ?} negative log-likelihood is given up to a constant independent of
$A_i$ and $S$ by:
\begin{align}
  \loss = \frac1{n} \sum_i \|A_i S - X_i \|^2 = \frac1{n} \big(\| S \|^2 -2 \dotp{ A_i S}{ X_i } + \| X_i \|^2 \big)
  \label{eq:detsrmloss}
\end{align}
$\loss$ is optimized by performing alternate minimization on $(A_1 \dots A_m)$
and $S$.
%
Note that the hyper-parameter $\sigma$ does not have an influence on
the loss and can therefore be ignored.

@@

From \eqref{eq:detsrmloss}, minimizing $\loss$ with respect to $A_i$ is
equivalent to maximizing $\dotp{ A_i}{ X_i S^{\top} }$ and therefore we
have:
\begin{equation}
  A_i \leftarrow  \Pcal \left(\frac1{n}X_i S^{\top} \right)
  \label{eq:detsrm:Aiupdate}
\end{equation}
where $\Pcal$ is the projection on the Stiefel manifold: $\Pcal(M) = M
(M^{\top}M)^{-\frac12}$.
%

The complexity of Deterministic SRM is in $\bigO(\mathrm{n_{iter}} mpvn)$ where
$n$ is the number of samples and $\mathrm{n_{iter}}$ the number of iterations.
%
We monitor the convergence by looking at the $\ell_{\infty}$ norm of the
gradient.
%
Note that we can monitor the gradient without any increase in complexity.
%
Indeed, after the updates with respect to each mixing matrix have been
carried out, only the gradient with respect to $S$ remains \textbf{XXX WDYM ?}: $\sum_i
(S - A_i^{\top}\xb_i)$.
%
The algorithm is stopped when the
gradient falls below a chosen tolerance.
%

@@

In~\cite{chen2015reduced} and~\cite{anderson2016enabling}, $\Sigma_s$ is only assumed to be definite positive.
%
As already highlighted in introduction, this causes the model to be
unidentifiable (see~\cite{richard2020modeling} Appendix D for a proof). Enforcing a diagonal $\Sigma_s$ ensures identifiability, provided that the diagonal values are different.
%
So we assume here that $\Sigma_s$ is diagonal (and refer the interested reader
to~\cite{chen2015reduced} and~\cite{anderson2016enabling} for the original formulation of Probabilistic SRM
without the diagonal constraint)
%

@@

\subsection{Reducing the computational burden by the use of spatial decompositions}

\textbf{XXX Sorry but we can't call $U$ and atlas: it does not define regions, it has no sign constraints, it is not sparse etc. It is actually closer to a random projection matrix than two an atlas ! So I propose to systematically remove the term atlas, and use spatial decompositions}

SRM algorithms use different sets of parameters $\theta$ to represent the data.
%
In deterministic SRM $\theta = \left((A_i)_{i=1}^m, S \right)$ where $(A_i)_{i=1}^m$ are the
mixing matrices and $S$ is the shared response, while in probabilistic SRM $\theta
= \left( (A_i)_{i=1}^m, \Sigma_s, (\sigma_i)_{i=1}^m \right) $ where $(A_i)_{i=1}^m$ are the
mixing matrices, $(\sigma_i)_{i=1}^m$ the noise levels and $\Sigma_s$ the
components variance.
%

@@

In FastSRM we consider a set of view-specific atlases $U_i \in \RR^{v \times r}$ such that
$U_i^{\top}U_i = I_r$ where $r$ is the number of regions in the atlas.
%
Data are reduced using $\zb_i = U_i^{\top} \xb_i$ and an SRM algorithm is applied
on data $\zb_i$ yielding parameters $\theta'$.
%
Figure~\ref{fig:srm:conceptual} illustrates this process.
%

@@

\subsection{A class of optimal spatial decompositions}
\label{sec:optimal_atlas}
In principle, FastSRM can be used with any atlas.
%
However, in general, working with reduced data induces a loss of information
that can be minimized if the atlas is carefully chosen.
%
In any case, there is little hope to recover the parameters that would have been
obtained from SRM from the parameters of FastSRM.
%
Yet, we show that there exists an optimal atlas in the
sense that SRM and FastSRM yield the same results.

Let us consider $\xb_i = U_{\xb_i} \zb_i$
%
% \bt{But any fill-rank orthogonal basis would work equally well (i.e. any
%   orthogonal transform of the PCA basis)}
%
a PCA of $\xb_i$ with the maximum
number of components.
%
As the number of
samples $n$ is lower than the number of features, $U_{\xb_i} \in \RR^{v \times
  n}$ and $\zb_i \in \RR^{n}$.
%
We also have $U_{\xb_i}^{\top}U_{\xb_i} = I$.
%
Therefore $U_{\xb_i}$ constitutes a possible choice of subject specific atlas.
%
As the next property shows, $U_{\xb_i}$ is an optimal atlas for deterministic
FastSRM. Note that this optimal atlas is not unique: any orthogonal matrix with
the same span as the columns of $X_i$ would also work.
%

@@

In the case of probabilistic SRM we can obtain very similar results.
%
However the
algorithm applied on reduced data needs to be slightly modified.
%

We have the following result:
\begin{proposition}[Optimal spatial decomposition for probabilistic FastSRM]
  Let $(A_i)_i, \sigma_i, \Sigma_s$ be the solution obtained by probabilistic SRM on data
  $\xb_i$ and $(A_i')_i, \sigma_i', \Sigma_s'$ the solution obtained by ProbSRM on
  data $\zb_i = U_{\xb_i}^{\top} \xb_i$ but where updates:
  \begin{align}
    \sigma_i^2 \leftarrow \frac1{t} (\EE[\| \zb_i - A_i \EE[\sbb|\zb]\|^2] + \| \diag(\VV[\sbb | \zb]) \|^2)
  \end{align}
  are replaced by updates
  \begin{align}
    \sigma_i^2 \leftarrow \frac1{v} (\EE[\| \zb_i - A_i \EE[\sbb|\zb]\|^2] + \| \diag(\VV[\sbb | \zb]) \|^2)
  \end{align}
  .
  %
  Then $A_i = U_{\xb_i}A_i'$, $\sigma_i =
  \sigma_i'$ and $\Sigma_s = \Sigma_s'$.
  %

@@

A complexity analysis shows that finding an optimal spatial decomposition becomes the limiting
step of the pipeline.
%
Even with fast implementations, subject specific PCA
is costly.
%
However FastSRM only works on $\zb_i$ so we do not need to know the value of
$U_{\xb_i}$.
%
In practice, we observe data $X_i \in \RR^{v \times n}$ and we want to get $Z_i
\in \RR^{n \times n}$ such that $X_i = U_{\xb_i} Z_i$.
%
This can be done by performing an
SVD of $X_i^{\top} X_i$ yielding $X_i^{\top}X_i= V_i D_i V_i^{\top}$ and setting
$Z_i = D_i^{\frac12} V_i^{\top}$.
%
Although computing the product $X_i^{\top} X_i$ has time complexity
$\bigO(vt^2)$ (the same as a PCA) the constant in the $\bigO$ is exactly one so
this operation costs a lot less than a PCA
on full data.
%
When estimates of the mixing matrices are needed, they can be obtained by
applying equation~\eqref{eq:fastdetsrm:Aiupdate} in the deterministic SRM case and
equation~\eqref{eq:fastprobsrm:Aiupdate} in the probabilistic SRM case which only costs
$\bigO(mvp^2)$.
%
In practice the cost of the matrix products $X_i^{\top} X_i$ is often still the
limiting step of the pipeline (this depends on the number of iterations) but as
we show next, it is much more efficient than performing SRM on
the full data.
%
Note that if memory allows it, these matrix products can be
computed in parallel.
%

\section{Related work}
The implementation of SRM proposed in~\cite{chen2015reduced} is quadratic in the
number of features which prevent its application to full brain data.
In~\cite{anderson2016enabling}, this issue is addressed by using the
inversion lemma to remove the quadratic dependence. Their implementation is now the most widely used one.
In our work, we propose to reduce further the computation time and memory usage
compared to the implementation in~\cite{anderson2016enabling} by the use of
optimal spatial decompositions.

@@

We benchmark deterministic SRM, probabilistic SRM and their FastSRM counterparts in terms of fitting time and performance.
%
Note that in this section, the identifiable implementation of deterministic SRM
and probabilistic SRM described in section~\ref{sec:probabilisticsrm}
and~\ref{sec:deterministicsrm} is used. 
%
Algorithms are designated by the atlas they use and therefore SRM algorithms are referred to
as \emph{None} because no atlas is used and FastSRM algorithms will have the
label \emph{Optimal}.
%
Note that it would be possible to use FastSRM with sub-optimal
atlases (there exists a wide variety of brain atlases
~\cite{schaefer2017local, bellec2010multi, mensch2018extracting}) but
using them does not bring any guarantee that the performance is the same as SRM.
%

@@

\subsection{Identifiability increases stability}
Assuming that the data follow the model, identifiability ensures that the
problem is well posed since the solutions are well-characterized. Note however that SRM
remains non-convex so there is no guarantee that the optimal solution is found.
%
% \bt{I find this a bit ambiguous: you give te impression that you gurantee that the optimal solution is used. Rephrase.}
%
In real fMRI data, the model cannot be expected to hold perfectly, but we can
hope for greater stability in the parameters recovered than if an unidentifiable
model is used.

To measure the stability of the shared components obtained from the Sherlock
dataset, we divide the subjects into two roughly equal groups and extract the
common components in each group.
%
The components are then matched using the Hungarian algorithm and the stability
measure is obtained from the average correlation of the matched components.
%
The procedure is repeated 9 times using the Brainiak implementation (that is not
identifiable since the shared components covariance is unconstrained) and our
FastSRM implementation.
%
We plot the results as a scatter plot in Figure~\ref{exp:identifiability}, where, for each repetition , the x-axis indicates the stability measure obtained with Brainiak's implementation
 and the y-axis the stability measure
obtained with FastSRM (identifiable since the shared components covariance is diagonal).
%
We see that introducing a diagonal source covariance improves stability.
%

@@

We show on synthetic and real data that FastSRM is much faster and more memory
efficient that implementations not using an optimal atlas.
Furthermore, FastSRM is identifiable. On real data, we show that compared to
Brainiak's implementation, that is not identifiable, FastSRM provides more
stable estimates of the shared components.

%
FastSRM inherits from all the weaknesses of SRM including the fact that
mixing matrices are assumed to be orthogonal, which is rather unrealistic. It
remains to be seen whether dimension reduction and learning of the shared
components could be done jointly and efficiently without assuming orthogonal
mixing matrices.
%
Our optimal atlas approach shows that SRM-like methods do not need the
full data to provide accurate estimates of shared components. We believe such
insights may guide the design of future methods.
%

Lastly, FastSRM is useful when the number of features is much larger than the
number of samples. This setting is common in fMRI but less common in MEG/EEG where
the number of samples is usually much larger than the number of features. Even
in this case, it is possible to considerably speed up SRM. Indeed, SRM only
depends on second order statistics.
These statistics can be pre-computed making the rest of the
pipeline insensitive to the number of samples in the dataset. An implementation
of this simple idea would make SRM easily applicable to other modalities.
We leave it to future work.
% \bt{Add a paragraph on implementation, sofwtare quality, documentation, examples, availability...}
% \bt{Discuss the use for other brain imaging modelities (MEEG) if relevant ?}
